---
title: "Residuals and Diagnostics in Generalized Linear Models"
author: "Rebecca Gill"
date: "`r Sys.Date()`"
header-includes:
  - \usepackage{amsmath}
output:
  bookdown::html_document2:
    number_sections: FALSE
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    code_folding: hide
    self_contained: true
    base_format: rmarkdown::html_document
bibliography: site_refs.bib
csl: gill_apsr.csl
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../docs"
    )
  })
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  cache     = FALSE,
  fig.align = "center"
)

options(scipen = 999)
options(digits = 4)

```

```{css, echo=FALSE}

/* This code defines the question div */

.question {
  border: 1px solid #cccccc;
  background-color: #f4faff;
  padding: 1em;
  border-radius: 5px;
  margin: 1em 0;
}

mark {
  background-color: LemonChiffon;
  color: black;
}

table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
  max-width: 80%;
  white-space: nowrap;
}

```

```{r libs}

P <- c(
  "knitr",          # rendering
  "kableExtra",     # table formatting
  "statmod",        # randomized quantile residuals (analytical)
  "DHARMa",         # simulation-based quantile residuals
  "performance",    # check_model(), check_residuals()
  "ggResidpanel",   # resid_panel(), resid_compare()
  "lme4",           # glmer() for GLMM examples
  "MASS",           # glm.nb() for negative binomial
  "patchwork",      # combining ggplots
  "tidyverse"       # wrangling and ggplot2 — load last
)

for (pkg in P) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
  }
  library(pkg, character.only = TRUE)
}

rm(P, pkg)  # clean up the mess

```

```{r sims}

set.seed(1975)   # for reproducibility

# ── Simulation 1: Poisson data with a hidden nonlinear term ───────────────────
# true data-generating process includes x1^2, but we'll fit the wrong model
# this comes from a teaching example in Craig (2024), adapted here

n.poi <- 400                                       # number of observations

x1 <- rnorm(n.poi, mean = 0, sd = 1.5)            # continuous predictor 1
x2 <- rnorm(n.poi, mean = 0, sd = 1.5)            # continuous predictor 2

eta.true <- 2.0 + 0.2*x1 + 0.1*x1^2 - 0.15*x2   # true linear predictor (has quadratic)
y.count  <- rpois(n.poi, lambda = exp(eta.true))   # Poisson counts

poi.df <- data.frame(y = y.count, x1 = x1, x2 = x2)  # bundle into data frame

# misspecified model: omits the x1^2 term
poi.wrong <- glm(y ~ x1 + x2,
                 family = poisson(link = "log"),
                 data   = poi.df)

# correctly specified model: includes x1^2
poi.right <- glm(y ~ x1 + I(x1^2) + x2,
                 family = poisson(link = "log"),
                 data   = poi.df)

# ── Simulation 2: Binary logistic data ───────────────────────────────────────
# well-specified binary model; used to show the limits of standard residuals

n.bin <- 600                                           # number of observations
x.bin <- rnorm(n.bin)                                  # single predictor
p.bin <- plogis(-0.3 + 1.5 * x.bin)                   # true probabilities
y.bin <- rbinom(n.bin, size = 1, prob = p.bin)         # binary outcome

bin.df <- data.frame(y = y.bin, x = x.bin)

logit.mod <- glm(y ~ x,
                 family    = binomial(link = "logit"),
                 data      = bin.df,
                 na.action = na.omit)

# ── Simulation 3: Clustered binary data for GLMM ──────────────────────────────
# students nested within schools; random intercept by school

n.groups  <- 25                                    # number of schools
n.per     <- 30                                    # students per school
n.total   <- n.groups * n.per                      # total observations

school.id <- rep(seq_len(n.groups), each = n.per)  # school membership
u.j       <- rnorm(n.groups, mean = 0, sd = 0.9)   # true random intercepts
u.i       <- u.j[school.id]                         # expand to observation level

x.clust   <- rnorm(n.total)                          # student-level predictor
eta.clust <- -0.4 + 0.9 * x.clust + u.i             # linear predictor
y.clust   <- rbinom(n.total, 1, plogis(eta.clust))  # binary outcome

glmm.df <- data.frame(
  y      = y.clust,
  x      = x.clust,
  school = factor(school.id)
)

# fit the GLMM
glmm.mod <- lme4::glmer(
  y ~ x + (1 | school),
  data   = glmm.df,
  family = binomial(link = "logit")
)

# ── Simulation 4: Negative binomial (for NB-specific quantile residuals) ───────
n.nb   <- 400
x.nb   <- rnorm(n.nb)
mu.nb  <- exp(1.5 + 0.8 * x.nb)                    # conditional mean
y.nb   <- rnegbin(n.nb, mu = mu.nb, theta = 2)     # NB counts (theta = precision)

nb.df  <- data.frame(y = y.nb, x = x.nb)

nb.mod <- MASS::glm.nb(y ~ x, data = nb.df)

```

---

## Introduction

If you've spent any time with linear regression, you know that residuals are the workhorses of diagnostics. You plot them, you stare at them, you look for patterns that shouldn't be there. The process feels almost meditative by the time you've done it enough times.

So here's a question worth asking before we go any further: *what exactly is a residual?*

In a linear model, the answer seems almost too simple to state. A residual is the difference between what you observed and what the model predicted:

$$r_i = y_i - \hat{y}_i$$

That's it. And because OLS models have normally distributed errors by assumption, and because the residuals are consistent estimators of those errors, the residuals themselves should look approximately normal. Their distribution is familiar. You know what "good" looks like.

Generalized linear models break all of this. The response variable is no longer assumed to be normally distributed — it might be binary, or a count, or ordinal. The conditional mean is no longer a simple linear function of the predictors; it's a transformation of one. And the variance of the outcome is no longer constant — it depends on the mean in a way that's specific to the distributional family. All of the comfortable intuitions you've built up about residuals need to be revised.

This page works through what residuals mean in the GLM context, why the most natural choices turn out to have serious limitations, and what current best practice looks like. We'll end up at **randomized quantile residuals** [@dunnSmyth1996] and their simulation-based cousins from the `DHARMa` package [@DHARMa] — and you'll understand *why* these approaches work, not just *how* to run them.

---

## Part I: What We Mean When We Say "Error" in a GLM

Before we look at residuals, it's worth being precise about what "error" means in a GLM, because it's subtler than in OLS.

In a standard linear model, the error is an explicit additive term:

$$Y_i = \mu_i + \varepsilon_i, \qquad \varepsilon_i \sim N(0, \sigma^2)$$

The error $\varepsilon_i$ lives alongside the mean. We estimate it directly as a residual, and because it's assumed normal, normal-based plots and tests apply immediately.

In a GLM, there is no separate additive error term. Instead, the model says:

$$Y_i \sim \text{EFD}(\mu_i, \phi)$$

where EFD is some exponential family distribution (Poisson, Binomial, Gamma, etc.) with conditional mean $\mu_i = g^{-1}(\mathbf{x}_i^\top\boldsymbol{\beta})$ and dispersion parameter $\phi$.[^1] The randomness in $Y_i$ comes entirely from the assumed distribution — it's baked into the distributional family, not tacked on as a separate error term. 

[^1]: The identity link gives us the familiar Gaussian GLM, which is equivalent to OLS. The log link gives us Poisson or negative binomial. The logit link gives us binary logistic regression. Each link function transforms the mean so that the linear predictor can range over the real line while the predicted mean stays in a sensible range for the outcome.

This has a practical consequence: we can't define a residual simply as $y_i - \hat{y}_i$ and expect it to behave nicely, because $y_i$ and $\hat{y}_i$ come from a distribution that may be highly skewed, bounded, or discrete. The difference between an observed binary outcome (0 or 1) and a predicted probability (say, 0.73) isn't normally distributed, no matter how well-specified the model is. It's a messy thing to look at.

> The residuals from a GLM don't estimate additive errors because GLMs don't have additive errors. What we call "residuals" in the GLM context are various summaries of the discrepancy between observed and fitted values — but we have to be careful about which summary we use and what we expect it to look like.

---

## Part II: A Taxonomy of GLM Residuals

Because there's no single natural residual for GLMs, statisticians have developed several alternatives, each designed to capture a different aspect of the model-data discrepancy. The following taxonomy draws on @craig_glm_diag and @fox2016. For each type, I'll show the formula, the R code, and what you'd use it for.

### Response Residuals

The response residual is the simplest: just the difference between the observed and fitted values, on the original scale of the response:

$$r_{i,R} = y_i - \hat{\mu}_i$$

```{r resid-response}

# response residuals are easy — they're the default for lm() but not for glm()
r.response <- residuals(poi.wrong, type = "response")

# quick look
summary(r.response)

```

These are intuitive to think about — the residual is in the same units as $Y$ — but they have a critical flaw: their variance is not constant. For a Poisson model, the variance of $Y$ is equal to its mean ($\text{Var}(Y) = \mu$), so observations with larger fitted values will naturally have more variable residuals. This makes plots of response residuals against fitted values difficult to interpret, because what looks like a funnel shape might just be a correct model respecting the Poisson variance structure rather than evidence of misfit.

### Working Residuals

GLMs are typically estimated via **iteratively reweighted least squares (IRWLS)**, and the working residual is the residual from the final iteration of that process — the difference between the observed and fitted values on the *link scale*:

$$r_{i,W} = Z_i - \hat{\eta}_i$$

where $Z_i$ is the "working response" (the adjusted dependent variable in the last IRWLS step) and $\hat{\eta}_i$ is the fitted linear predictor. Adding $x_{ij}\hat{\beta}_j$ to the working residual gives you the **partial residual** for variable $j$, which is the basis of component-plus-residual plots:

```{r resid-working}

# working residuals
r.working <- residuals(poi.wrong, type = "working")

# partial residuals (one per predictor) — used for linearity checks
r.partial.x1 <- residuals(poi.wrong, type = "partial")[, "x1"]
r.partial.x2 <- residuals(poi.wrong, type = "partial")[, "x2"]

```

Working residuals are the basis of many diagnostic plots in software, but they're on the link scale (e.g., the log scale for a Poisson model), which makes them less interpretable for general audiences.

### Pearson Residuals

The Pearson residual standardizes the response residual by the model-implied standard deviation of $Y_i$:

$$r_{i,P} = \frac{y_i - \hat{\mu}_i}{\sqrt{V(\hat{\mu}_i)}}$$

where $V(\hat{\mu}_i)$ is the variance function of the distribution evaluated at $\hat{\mu}_i$. For a Poisson model, $V(\mu) = \mu$, so the Pearson residual divides by $\sqrt{\hat{\mu}_i}$. For a binomial model with $n_i$ trials, $V(\mu) = \mu(1-\mu)/n_i$.

The sum of squared Pearson residuals gives the Pearson $\chi^2$ statistic: $\sum r_{i,P}^2 = X^2$.

```{r resid-pearson}

# pearson residuals
r.pearson <- residuals(poi.wrong, type = "pearson")

# the sum of squared pearson residuals gives the Pearson chi-squared statistic
x2.stat <- sum(r.pearson^2)
cat("Pearson chi-squared:", round(x2.stat, 2), "\n")
cat("Degrees of freedom:", df.residual(poi.wrong), "\n")
cat("Ratio (should be ~1 if equidispersed):", round(x2.stat / df.residual(poi.wrong), 3), "\n")

```

Pearson residuals have approximately constant variance if the variance function is correctly specified, making them more useful for heteroscedasticity checks than response residuals. The standardized version goes one step further.

### Standardized Pearson Residuals

The standardized Pearson residual adjusts for leverage ($h_{ii}$, the diagonal of the GLM hat matrix) and the dispersion parameter $\hat{\phi}$:

$$r_{i,SP} = \frac{y_i - \hat{\mu}_i}{\sqrt{\hat{\phi}\, V(\hat{\mu}_i)(1 - h_{ii})}}$$

This is the GLM analogue of the studentized residual from OLS. It should have approximately constant variance if the model is correctly specified, making it the best choice among the traditional residual types for detecting variance misspecification. In R, it's available through the `boot` package:[^2]

[^2]: The `boot::glm.diag()` function is the old-school workhorse for GLM diagnostics. It returns a list with components `rp` (standardized Pearson), `rd` (standardized deviance), `res` (jackknife), `cook` (Cook's distance), and `h` (leverage). It works on any `glm` object. The individual functions like `cooks.distance()` and `hatvalues()` from base R also work on `glm` objects directly and are often easier to work with.

```{r resid-std-pearson}

library(boot)   # for glm.diag()

# glm.diag() returns several diagnostics at once
poi.diag <- boot::glm.diag(poi.wrong)

r.std.pearson <- poi.diag$rp   # standardized Pearson residuals

```

### Deviance Residuals

The deviance residual is based on each observation's contribution to the model's overall deviance $D$. If $d_i$ is observation $i$'s contribution to the deviance, then:

$$r_{i,D} = \text{sign}(y_i - \hat{\mu}_i)\sqrt{d_i}$$

The sum of squared deviance residuals equals the total residual deviance: $\sum r_{i,D}^2 = D(\hat{\boldsymbol{\beta}})$.

Deviance residuals tend to be less skewed than Pearson residuals for most GLM families, which makes them somewhat more useful for outlier detection — *except for binary outcomes*, where even deviance residuals are badly non-normal (more on this below).

```{r resid-deviance}

# deviance residuals — the default for glm objects
r.deviance <- residuals(poi.wrong, type = "deviance")

# standardized version via glm.diag()
r.std.deviance <- poi.diag$rd   # standardized deviance residuals

# jackknife residuals (approximated)
r.jackknife <- poi.diag$res

```

The **standardized deviance residual** adjusts for leverage the same way the standardized Pearson residual does, and the **jackknife (deleted studentized) residual** approximates the change in deviance when observation $i$ is removed. The jackknife residual is generally a good all-purpose choice among the traditional options [@craig_glm_diag].

### Summary Table

```{r resid-summary-table, results='asis'}

resid.tbl <- data.frame(
  Type = c(
    "Response",
    "Working",
    "Pearson",
    "Std. Pearson",
    "Deviance",
    "Std. Deviance",
    "Jackknife"
  ),
  Formula_sketch = c(
    "$y_i - \\hat{\\mu}_i$",
    "$Z_i - \\hat{\\eta}_i$",
    "$(y_i - \\hat{\\mu}_i)/\\sqrt{V(\\hat{\\mu}_i)}$",
    "Pearson / $\\sqrt{\\hat{\\phi}(1-h_{ii})}$",
    "$\\text{sign}(y_i-\\hat{\\mu}_i)\\sqrt{d_i}$",
    "Deviance / $\\sqrt{\\hat{\\phi}(1-h_{ii})}$",
    "Approx. deleted studentized"
  ),
  R_code = c(
    '`residuals(m, type="response")`',
    '`residuals(m, type="working")`',
    '`residuals(m, type="pearson")`',
    '`glm.diag(m)$rp`',
    '`residuals(m, type="deviance")`',
    '`glm.diag(m)$rd`',
    '`glm.diag(m)$res`'
  ),
  Best_for = c(
    "Intuition; same units as Y",
    "Linearity via partial residual plots",
    "Variance misspecification; Pearson $\\chi^2$",
    "Outlier detection; constant variance",
    "Outlier detection (less skewed than Pearson)",
    "Outlier detection; constant variance",
    "General-purpose; best of the traditional options"
  )
)

kable(resid.tbl,
      col.names = c("Type", "Formula (sketch)", "R code", "Best for"),
      caption   = "Traditional GLM residual types",
      escape    = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width        = FALSE
  )

```

---

## Part III: Why None of Them Are Quite Right

Here's the uncomfortable truth that the table above doesn't fully communicate: **for non-Gaussian GLMs, all of the residual types above have properties that make them hard to use reliably.**

The core problem is that they all try to use continuous, approximately-normal approximations to summarize discrepancies in a non-normal, sometimes discrete distribution. The approximations work reasonably well in some cases (e.g., large-count Poisson data where the normal approximation to the Poisson kicks in), but they break down in exactly the cases where you most need reliable diagnostics.

Consider the binary logistic regression case. With a binary outcome ($y_i \in \{0, 1\}$), the response residual takes only two possible values for any given fitted probability $\hat{p}_i$: either $1 - \hat{p}_i$ (when $y_i = 1$) or $-\hat{p}_i$ (when $y_i = 0$). No matter how good your model is, a QQ plot of these residuals will look terrible — because the residuals can't possibly be normal when the outcome can only be 0 or 1.

```{r binary-resid-problem}
#| fig.cap: "QQ plots of three residual types for a well-specified binary logistic model. The left panel shows the theoretical problem: even a correctly specified model produces non-normal response residuals. Deviance residuals (center) are better but still non-normal. Pearson residuals (right) show the same issue."

par(mfrow = c(1, 3))   # three-panel plot

# response residuals — unmistakably bad even under a good model
qqnorm(residuals(logit.mod, type = "response"),
       main = "Response Residuals",
       pch  = 20, col = "grey40", cex = 0.6)
qqline(residuals(logit.mod, type = "response"), col = "red")

# deviance residuals — better, but still problematic for binary data
qqnorm(residuals(logit.mod, type = "deviance"),
       main = "Deviance Residuals",
       pch  = 20, col = "grey40", cex = 0.6)
qqline(residuals(logit.mod, type = "deviance"), col = "red")

# pearson residuals
qqnorm(residuals(logit.mod, type = "pearson"),
       main = "Pearson Residuals",
       pch  = 20, col = "grey40", cex = 0.6)
qqline(residuals(logit.mod, type = "pearson"), col = "red")

```

::: question
Are you telling me that there's *no way* to use these residuals to diagnose a binary logistic regression?
:::

Not exactly. They're not useless — Pearson and deviance residuals can still flag outliers and influential observations, and working/partial residuals are still useful for linearity checks. But for assessing the overall distribution of residuals (via QQ plots) or for detecting subtle misspecification, the traditional residuals are unreliable for binary outcomes and for any other discrete distribution.

The issue persists, in a milder form, for count models too. Even for Poisson data, the discrete nature of the outcome means that residuals of all traditional types can show non-normality that's entirely due to the discreteness of the distribution rather than to model misfit. This makes it hard to separate signal from noise in QQ plots.

> <mark>The standard QQ plot of deviance or Pearson residuals is appropriate for Gaussian GLMs but should be interpreted with great caution — and often avoided entirely — for binary, count, and other discrete-outcome GLMs.</mark>

---

## Part IV: Randomized Quantile Residuals

The solution proposed by @dunnSmyth1996 is elegant. The key insight comes from the **probability integral transform**: if $Y$ is a continuous random variable with CDF $F$, then $F(Y) \sim \text{Uniform}(0, 1)$ regardless of the distribution of $Y$. And if $U \sim \text{Uniform}(0,1)$, then $\Phi^{-1}(U) \sim N(0,1)$ (where $\Phi^{-1}$ is the standard normal quantile function). So:

$$r_{i,q} = \Phi^{-1}(F(y_i \mid \hat{\mu}_i, \hat{\phi}))$$

If the model is correctly specified, the CDF evaluated at the observed value $y_i$ should look like a draw from a Uniform(0,1) distribution, and therefore the quantile-transformed version $r_{i,q}$ should look like a draw from $N(0,1)$. <mark>Under a correctly specified model, randomized quantile residuals are exactly standard normal — not approximately, not asymptotically, but exactly (for continuous distributions).</mark>

::: question
That's for continuous distributions. What about binary outcomes and count data, where $F$ is a step function?
:::

Right, and this is where the "randomized" part comes in. For discrete distributions, the probability integral transform doesn't produce a uniform distribution because the CDF is a step function — it jumps from value to value rather than varying continuously. The fix is to **randomize within the jump**:

Let $a_i = F(\max(0, y_i - 1) \mid \hat{\mu}_i, \hat{\phi})$ and $b_i = F(y_i \mid \hat{\mu}_i, \hat{\phi})$. Then draw $u_i \sim \text{Uniform}(a_i, b_i)$ and define:

$$r_{i,q} = \Phi^{-1}(u_i)$$

This single random draw fills in the gap at the step, making the distribution of $r_{i,q}$ continuous even when $Y$ is discrete. Under a correctly specified model, the $r_{i,q}$ are exactly $N(0,1)$ — full stop.

The only catch is that because $u_i$ is random, the specific values of $r_{i,q}$ will differ slightly from one run to the next. The *pattern* in the residuals — whether they look normal, whether there are outliers, whether there's structure — stays the same. Only the exact positions of individual points jitter slightly.

```{r qr-statmod}
#| fig.cap: "Quantile residuals (via `statmod`) for the misspecified Poisson model (left) and the correctly specified model (right). The QQ plots should show straight lines under a good model. The pattern of deviation on the left reveals the missing nonlinear term."

# statmod provides analytical quantile residuals for common GLM families
# qresid() is the general function; qres.pois() is family-specific

r.qr.wrong <- statmod::qres.pois(poi.wrong)   # wrong model
r.qr.right <- statmod::qres.pois(poi.right)   # correct model

par(mfrow = c(1, 2))

# QQ plot for misspecified model
qqnorm(r.qr.wrong,
       main = "Misspecified Model\n(missing x1²)",
       pch  = 20, col = "grey40", cex = 0.6)
qqline(r.qr.wrong, col = "red")

# QQ plot for correctly specified model
qqnorm(r.qr.right,
       main = "Correctly Specified Model",
       pch  = 20, col = "grey40", cex = 0.6)
qqline(r.qr.right, col = "red")

```

The deviation in the left panel is unmistakable. The correctly specified model (right) produces residuals that track the reference line closely — which is exactly what we're looking for.

The `statmod` package [@statmod] provides family-specific functions: `qres.pois()` for Poisson, `qres.nb()` for negative binomial, `qres.binom()` for binomial. A general `qresid()` function handles model objects from many families automatically:

```{r qr-general}
#| fig.cap: "Quantile residual plot: residuals vs. linear predictor for the misspecified Poisson model. The curved pattern reveals the missing nonlinear term in x1."

# general-purpose quantile residuals (works on glm objects)
r.qr <- statmod::qresid(poi.wrong)

# plot against the linear predictor (on the link scale)
plot(
  predict(poi.wrong),         # fitted linear predictor (log scale)
  r.qr,
  xlab = "Fitted Linear Predictor",
  ylab = "Quantile Residual",
  main = "Quantile Residuals vs. Fitted Values",
  pch  = 20, col = "grey40", cex = 0.6
)
abline(h = 0, col = "red", lty = 2)
lines(lowess(predict(poi.wrong), r.qr), col = "steelblue", lwd = 2)

```

You can also plot quantile residuals against individual predictors to locate the source of misfit:

```{r qr-by-predictor}
#| fig.cap: "Quantile residuals vs. x1 (left) and x2 (right) for the misspecified model. The curved pattern in the left panel points directly to the omitted x1² term; the right panel is clean."

par(mfrow = c(1, 2))

# against x1 — should show pattern from the missing quadratic
plot(x1, r.qr.wrong,
     xlab = "x1", ylab = "Quantile Residual",
     main = "QR vs. x1 (misspecified)",
     pch  = 20, col = "grey40", cex = 0.6)
abline(h = 0, col = "red", lty = 2)
lines(lowess(x1, r.qr.wrong), col = "steelblue", lwd = 2)

# against x2 — should look clean (x2 is correctly modeled)
plot(x2, r.qr.wrong,
     xlab = "x2", ylab = "Quantile Residual",
     main = "QR vs. x2 (misspecified)",
     pch  = 20, col = "grey40", cex = 0.6)
abline(h = 0, col = "red", lty = 2)
lines(lowess(x2, r.qr.wrong), col = "steelblue", lwd = 2)

```

The pattern in the x1 panel — a U-shape centered near zero — is the signature of a missing quadratic term. The x2 panel is noise around the zero line, as expected. This is exactly how you'd use quantile residuals in practice: first look at the QQ plot to see if there's misfit overall, then plot against predictors to locate it.

---

## Part V: DHARMa — Residuals by Simulation

The `statmod` approach to quantile residuals is elegant but requires knowing the analytical CDF of the outcome distribution. For complex models — particularly **GLMMs**, where the marginal distribution of $Y$ involves integrating over the random effects and has no closed-form CDF — this becomes impractical.

`DHARMa` [@DHARMa] takes a different approach: instead of computing the CDF analytically, it **simulates** many datasets from the fitted model and uses the empirical distribution of those simulations as a stand-in for the theoretical CDF.

The procedure:

1. Fit your model.
2. Simulate $S$ new response vectors from the fitted model, using the fitted parameter estimates.
3. For each observation $i$, compute the empirical quantile of $y_i$ in its simulated distribution: $\hat{F}(y_i) = \frac{1}{S}\sum_{s=1}^{S} \mathbf{1}[y_i^{(s)} \leq y_i]$.
4. Transform: $r_{i,q} = \Phi^{-1}(\hat{F}(y_i))$.

Under a correctly specified model, these simulated quantile residuals should be uniform on $(0,1)$ — equivalently, the $\Phi^{-1}$-transformed version should be $N(0,1)$. No analytical CDF required.

```{r dharma-basic, cache=TRUE}
#| fig.cap: "DHARMa residuals for the misspecified Poisson model (left) and the correctly specified model (right). The QQ plot plots simulated quantile residuals against the expected uniform distribution; the residual-vs-fitted plot checks for systematic patterns."

# simulate residuals from the misspecified model
# n = 500 simulations is enough for stable results; more = slower but smoother
sim.wrong <- DHARMa::simulateResiduals(poi.wrong, n = 500, plot = FALSE)
sim.right <- DHARMa::simulateResiduals(poi.right, n = 500, plot = FALSE)

par(mfrow = c(1, 2))

# DHARMa's built-in plot: QQ plot (left) + residual vs. rank of fitted (right)
plot(sim.wrong, main = "Misspecified Model")

```

```{r dharma-right, cache=TRUE}
#| fig.cap: "DHARMa residuals for the correctly specified Poisson model. Compare with the misspecified version above."

plot(sim.right, main = "Correctly Specified Model")

```

DHARMa's default plot shows two panels: a QQ plot of the simulated residuals against the expected uniform distribution, and a scatter plot of residuals against the rank of fitted values. Red asterisks in the QQ plot and colored bands in the residual plot indicate where significant deviations occur.

::: question
Wait — if the residuals are constructed from random simulations, won't the plots look a little different every time I run them?
:::

Yes, slightly — but the patterns are stable across runs for any reasonable number of simulations ($n \geq 250$ is a good floor; $n = 500$ is reliable for most purposes). The *specific* residual value for any given observation jitters a tiny bit from run to run (this is the same randomization step as in `statmod`), but the overall shape of the QQ plot, the presence of curvature in the residual-vs-fitted plot, and the significance of formal tests remain consistent. Think of each plot as a sample from the distribution of plots you'd get — the signal stays, the noise varies slightly.

### Formal Tests

DHARMa also provides hypothesis tests for specific types of misfit. These are useful for getting a second opinion when a plot is ambiguous.

```{r dharma-tests}

# test for uniformity of residuals (overall goodness of fit)
DHARMa::testUniformity(sim.wrong)

# test for dispersion (variance larger or smaller than model predicts)
DHARMa::testDispersion(sim.wrong)

# test for outliers (observations far outside the simulated distribution)
DHARMa::testOutliers(sim.wrong, type = "bootstrap", plot = FALSE)

```

The `testUniformity()` function runs a Kolmogorov-Smirnov test on the simulated residuals against the uniform distribution. `testDispersion()` checks whether the variance of the residuals is what the model predicts. `testOutliers()` asks whether there are more extreme observations than the model would expect by chance.

::: question
Should I report the results of these tests in a paper?
:::

Usually not — at least not as your primary evidence. These tests have well-known power issues: in large samples, trivial and substantively meaningless deviations can become significant; in small samples, important deviations can fly under the radar. The plots are your primary diagnostic tool. Use the tests as supplementary evidence when the visual is unclear, and always lead with a substantive discussion of what you see in the plots rather than a p-value.

### Plotting Against Predictors

Just as with `statmod`, you can and should plot DHARMa residuals against each predictor to locate the source of misfit:

```{r dharma-predictor}
#| fig.cap: "DHARMa residuals plotted against x1 for the misspecified model. The significant pattern (red banding) localizes the misfit to this predictor."

# plotResiduals() plots DHARMa residuals against a variable of your choice
DHARMa::plotResiduals(sim.wrong, form = poi.df$x1,
                      xlab = "x1",
                      main = "DHARMa Residuals vs. x1 (Misspecified Model)")

```

---

## Part VI: Practical Workflow Tools

The `statmod` and `DHARMa` approaches give you precise control over your diagnostic process. Two additional tools are worth knowing for a faster first-pass look at any model.

### `performance::check_model()`

The `performance` package [@performance] provides `check_model()`, which produces a comprehensive panel of diagnostics for a wide variety of model types. For non-Gaussian GLMs, it uses DHARMa under the hood when `residual_type = "simulated"` (the default for non-Gaussian models).

```{r check-model, eval=FALSE}

# one-stop diagnostic panel — best viewed interactively or in an HTML document
# the plot will include: posterior predictive check, QQ of residuals,
# residuals vs. fitted (linearity), homogeneity of variance, 
# influential observations (Cook's D), and multicollinearity (VIF)

performance::check_model(poi.wrong)

```

```{r check-residuals}

# check_residuals() gives a focused look at the DHARMa quantile residuals
# specifically: KS test + uniform QQ plot
performance::check_residuals(poi.wrong)

```

`check_residuals()` wraps DHARMa's `testUniformity()` and produces a clean plot suitable for including in a write-up. For any non-Gaussian GLM, this is a good first diagnostic step.

### `ggResidpanel`

The `ggResidpanel` package [@ggResidpanel] provides `resid_panel()` for a single model and `resid_compare()` for side-by-side comparison of multiple models. It works on `lm`, `glm`, `lmerMod`, and `glmerMod` objects.

```{r ggresidpanel}
#| fig.cap: "Side-by-side diagnostic panel comparing the misspecified (left) and correctly specified (right) Poisson models via ggResidpanel. The QQ plots and index plots are particularly useful here."

# compare two models side by side
# "default" panel: residual plot, QQ plot, index plot, histogram
ggResidpanel::resid_compare(
  models = list(poi.wrong, poi.right),
  plots  = "default",
  type   = "deviance",     # deviance residuals for Poisson
  smoother = TRUE,
  title.opt = TRUE,
  nrow   = 2
)

```

For a single model, `resid_panel()` gives a similar panel. Specifying `type = "pearson"` or `type = "deviance"` selects the residual type; the default for `glm` objects is deviance.

> One thing to be aware of: `ggResidpanel` uses traditional residual types (Pearson, deviance) and does not currently support randomized quantile residuals or DHARMa-based residuals. It's most useful for a quick look, and for Gaussian-family models where traditional residuals are well-behaved. For non-Gaussian models, treat `ggResidpanel` output with the cautions noted in Parts III and IV.

---

## Part VII: Two Levels of Error in GLMMs

Mixed models (GLMMs) add a new dimension to the residual problem: there are now *two* sources of random variation, not one. Observations within a group share a random effect, and then there's individual-level variation on top of that. When we talk about "residuals" for a GLMM, we need to be clear about which level we're talking about.

The model is:

$$g(\mu_{ij}) = \mathbf{x}_{ij}^\top\boldsymbol{\beta} + \mathbf{z}_{ij}^\top\mathbf{u}_j$$

where $\mathbf{u}_j \sim N(\mathbf{0}, \boldsymbol{\Sigma})$ are the group-level random effects. There are two types of "error" here:

- **Level-1 (individual) residuals**: $y_{ij} - \hat{\mu}_{ij}$ on the response scale, or the GLM analogue. These are what most diagnostic functions return by default.
- **Level-2 (group) residuals**: the estimated random effects $\hat{u}_j$ (BLUPs — Best Linear Unbiased Predictors). These capture how much each group's baseline deviates from the overall intercept.

Both levels deserve diagnostic attention.

### Level-1 Residuals: DHARMa for GLMMs

DHARMa handles GLMMs seamlessly — this is one of its main advantages over `statmod`, which requires an analytical CDF. Simulation from a fitted `glmer()` model automatically marginalizes over the random effects, producing valid quantile residuals at the observation level:

```{r dharma-glmm, cache=TRUE}
#| fig.cap: "DHARMa quantile residuals for the simulated GLMM. Under a correctly specified model, these should follow a uniform distribution."

# simulate residuals from the GLMM
# DHARMa's simulateResiduals() works on glmerMod objects
sim.glmm <- DHARMa::simulateResiduals(glmm.mod, n = 500, plot = FALSE)

# diagnostic plot
plot(sim.glmm, main = "GLMM Quantile Residuals (DHARMa)")

```

```{r dharma-glmm-tests}

# formal tests for the GLMM
DHARMa::testUniformity(sim.glmm)
DHARMa::testDispersion(sim.glmm)

```

### Level-2 Residuals: QQ Plot of Random Effects

The random effects $\hat{u}_j$ are assumed to be drawn from a normal distribution with mean zero. A QQ plot of the BLUPs is a useful check on this assumption — systematic deviations from normality might indicate that the random effects structure is misspecified, or that there are outlier groups.[^3]

[^3]: A QQ plot of BLUPs should be interpreted with care. BLUPs are shrunken toward zero by the mixed model's partial-pooling mechanism, especially when groups are small. This shrinkage tends to make the empirical distribution of BLUPs look more normal than the true random effects distribution actually is. Significant deviations in the tails of the BLUP QQ plot are meaningful; the center of the distribution will almost always look well-behaved regardless. See @bates2015 for discussion.

```{r ranef-qqplot}
#| fig.cap: "QQ plot of estimated random effects (BLUPs) for the school-level intercepts in the GLMM. Under correct specification of the random effects distribution, these should fall along the reference line."

# extract BLUPs (estimated random effects) from the GLMM
blup.vals <- ranef(glmm.mod)$school[, "(Intercept)"]

# QQ plot
qqnorm(blup.vals,
       main = "QQ Plot of School-Level Random Effects",
       pch  = 20, col = "steelblue")
qqline(blup.vals, col = "red")

```

`performance::check_model()` includes a QQ plot of random effects when run on a mixed model object — look for the panel labeled "Normality of Random Effects":

```{r performance-glmm, eval=FALSE}

# comprehensive diagnostic panel for the GLMM
# includes: QQ of residuals, QQ of random effects, residuals vs. fitted, etc.
performance::check_model(glmm.mod)

```

### Heterogeneity of Residual Variance Across Groups

One additional GLMM-specific check worth making is whether the level-1 residual variance is consistent across groups. If some groups have systematically larger residuals than others (after accounting for the random effect), this can indicate that the variance structure is wrong or that important group-level predictors are missing.

A simple visual check: plot the within-group mean absolute deviance residual against a measure of group size or group mean on $Y$.

```{r glmm-groupvar}
#| fig.cap: "Mean absolute deviance residual by school. If residuals are well-behaved, the group-level averages should be roughly uniform across schools."

# extract deviance residuals from the GLMM
glmm.df$resid <- residuals(glmm.mod, type = "deviance")

# compute within-group mean absolute residual
group.resid <- glmm.df %>%
  group_by(school) %>%
  summarise(
    mean.abs.resid = mean(abs(resid)),  # average absolute deviance residual
    n              = n()                # group size
  )

# plot
ggplot(group.resid, aes(x = reorder(school, mean.abs.resid),
                         y = mean.abs.resid)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  labs(
    x     = "School (sorted by mean absolute residual)",
    y     = "Mean Absolute Deviance Residual",
    title = "Within-Group Residual Variation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank())   # school IDs too small to read

```

Large systematic differences across groups — especially if they correlate with group size or group-level predictors — are a flag worth investigating.

---

## Summary: A Practical Hierarchy

When you're diagnosing a GLM or GLMM in practice, work through this sequence:[^4]

[^4]: This workflow synthesizes recommendations from @craig_glm_diag, @DHARMa, @performance, and @sscc_mm. It represents current best practice for diagnosing limited dependent variable models and mixed models, and deliberately moves away from the OLS-centric diagnostic habits (plotting raw deviance residuals, using `plot(model)`) that are common in older applied textbooks.

1. **Start with DHARMa.** Run `simulateResiduals()` and `plot()` on every non-Gaussian model as a first step. The QQ plot and residual-vs-fitted plot will catch most forms of misfit. For GLMMs, DHARMa is the most reliable tool available.

2. **For analytical quantile residuals** on single-level GLMs, `statmod::qresid()` or the family-specific functions are equivalent to DHARMa and require no simulation. Use them when you want deterministic (non-random) residuals for reproducibility.

3. **For a broader panel of diagnostics** — including Cook's D, leverage, VIF, and posterior predictive checks — use `performance::check_model()`. It calls DHARMa internally for non-Gaussian models.

4. **Use `ggResidpanel` for quick comparisons** across model specifications, keeping in mind that it uses traditional (not quantile) residuals.

5. **Check the random effects QQ plot** for every GLMM. Look at both the level-1 DHARMa residuals and the level-2 BLUP QQ plot.

6. **Do not** use QQ plots of raw deviance or Pearson residuals as your primary diagnostic for binary or count outcomes. They will almost always look non-normal even under a correctly specified model, making them uninformative or actively misleading.

---

## References
